---
title: "unicon"
output: rmarkdown::html_vignette
toc: true
toc_float: true
number_sections: true
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(unicon)
```

# Introduction

Different unit systems (most notably metric and imperial) are used interchangeably around the world to convey the same information. The International System of Units (SI, from the french *'Système international d'unités'*) aims to standardise unit measurements, allowing meaningful comparisons. However, the problem remains if data is collected using non-SI measurement systems. The {unicon} ('unit conversion') package provides a simple tool to enable users to quickly convert between different units of measurement using:

1. A database of base and derived units, including their various names, notations, and the SI equivalents chosen by authors to be most useful for their purposes, henceforth called standard reference point (SRP);
2. Functions to convert non-SRP units to SRP units and vice versa;
3. Functions to calculate additional units from existing data.
4. Relationships between different metrics to allow calculation of new metrics. E.g. Area * Length = Volume.

# Database of base and derived units

Three key mappings are created within the package:

1. Names and aliases for units. E.g. cm, centimetres, and centimeters are aliases for cm.
2. Calculations for converting non-SRP derived units into SRP derived units. E.g. multiplying cm/s by 0.01 to achieve m/s.
3. A list of what the recognised SRP unit is within the {unicon} package. E.g. SRP for length is m.

These data are used in the functions described below.

# Usage

## `unicon_full`

The core user-facing function of the `unicon` package is `unicon_full`. Use it like this:

```{r unicon_full-1}
raw_values <- c(54.21, 71.24, 55.81, 11.33, 70.59)
raw_units <- c("tonnes / ha", "tons per acre", "t/ha", "kg /Hectare", "g/m2")
unit_out <- "tonnes / ha"

unicon_full(
  value_in = raw_values,
  unit_in = raw_units,
  unit_out = unit_out,
  pull = TRUE
)
```

The function uses `unicon`'s extensive library of unit aliases to match the highly varied raw input units to their corresponding unit IDs, and subsequently converts them to the specified output unit.

A number of features are built into the function to help keep conversions safe and transparent. Firstly, the user may wish to see the full calculation thread which leads to this outcome, which can be achieved by setting the `pull` argument to FALSE. This yields a tibble, which we'll `glimpse`:

```{r unicon_full-2}
unicon_full(
  value_in = raw_values,
  unit_in = raw_units,
  unit_out = unit_out,
  pull = FALSE
) |>
  dplyr::glimpse()
```

This returns the following information, representing the calculation process in rough left --> right chronology:

* `unit_in` The user-supplied `unit_in` argument.
* `unit_out` The user-supplied `unit_out` argument.
* `alias_in` The derived alias associated to the `unit_in` argument, generated by stripping whitespace and converting all characters to lowercase.
* `alias_out` The derived alias associated to the `unit_out` argument, derived similarly.
* `id_in` The ID acquired for the `alias_in` variable.
* `id_srp` The ID associated with the "SRP" (standard reference point) unit for the type class of the input and output units.
* `id_out` The ID acquired for the `alias_out` variable.
* `error_in` Logical; were there errors in processing the input unit?
* `error_srp` Logical; were there errors in deriving the SRP unit?
* `error_out` Logical; were there errors in deriving the output unit?
* `value_in` The user-supplied `value_in` argument.
* `value_srp` The values as converted to their SRP (standard reference point) unit.
* `value_out` The values converted to the `unit_out` argument, if supplied. If not supplied, the standard reference point unit will be returned and the values will be identical to `value_srp`.

To understand more about the significance of the `srp` unit, read **A brief note on design and methodology**, below.

All the core input arguments are vectorised and will attempt to replicate if supplied as scalar values. This is helpful, as above, in specifying a single unit of output, or in converting vectors of consistent unit type (`unicon_full(my_values, "tonnes/ha", "g/m2")`). It's also possible to omit the `unit_out` argument altogether; if omitted, the function will return values in the associated standard reference unit and will message the user to indicate that this has been done.

Finally, although `unicon` has a very large library of unit aliases, it is probable that it will be passed units that it does not recognise and important that it handles these clearly:

```{r unicon_full-3, warning=TRUE}
unicon_full(
  value_in = raw_values,
  unit_in = c("tonnes / ha", "tons per acre", "t/ha", "kg /Hectare", "nonsense per fiction"),
  unit_out = unit_out,
  pull = TRUE
)
```

As previously, setting the `pull` argument to `FALSE` will yield better information as to the nature of the error.

```{r unicon_full-4, warning=TRUE}
unicon_full(
  value_in = raw_values,
  unit_in = c("tonnes / ha", "tons per acre", "t/ha", "kg /Hectare", "nonsense per fiction"),
  unit_out = "tonnes / ha",
  pull = FALSE
)
```

## `unicon_lite`

It is anticipated that a secondary use-case for the `unicon` functionality will be in a setting where mathematical unit conversion is required, but where the environment is tightly controlled to the point where recognition of thousands of aliases all pointing to the same models is surplus to requirements. The `unicon_lite` function is therefore designed to dispense with the alias-sifting that is handled automatically in `unicon_full`, and instead takes the unit ID as a direct argument for conversion:

```{r unicon_lite-1}
unicon_lite(
  value_in = raw_values,
  id_in = c("t__ha", "st__acre", "t__ha", "kg__ha", "g__m_2"),
  id_out = "t__ha",
  pull = TRUE
)
```

As with `unicon_full`, `unicon_lite` will return a full dataframe if required (`pull = TRUE`) and will warn if conversions fail.

## `unicon_catrel`

### Metric relationships 

The relationships between metrics that are included in this package are:

1.	**Area Density = Mass / Area**

2.  **Area = Volume / Length** where *Area* and *Volume* are also listed as a base units. Please calculate *Area* from 4. if unknown. 

3.	**Concentration = Moles / Volume**

4.  **Length = Area / Length** where lengths can be different values. Please note *Area* is also listed as a base unit.

5.  **Mass fraction = Mass / Mass**

6.	**Pressure = Force / Area**

7.	**Volume Density = Mass / Volume**

8.  **Volume fraction = Volume / Volume**

9.  **Speed = Length / Time**

### Examples

To calculate a new metric from existing data:

e.g. speed from length / time (if unit_out unspecified, will convert to SRP (in this example m/s))

```{r unicorn_catrel-1}
unicon_catrel(
  parent_1_unit_in = rep("miles", 6),
  parent_2_unit_in = rep("hour", 6),
  parent_1_value_in = c(1, 2, 3, 4, 5, 6),
  parent_2_value_in = c(9, 8, 7, 5, 4, 2),
  unit_out = rep("km/day", 6)
)
```

Note that two 'No output unit given. Converting all values to standard reference unit.' will appear, but ignore these when using the unicon_catrel function.

## `unicon_help`

To use the `unicon` functions, it may be useful to directly see what is recognised by the package. The internal package data is maintained in an optimal format for efficient computation, but the `unicon_help` function is designed to join and return this in the simplest format for interactive analysis:

```{r unicon_help}
unicon_help()
```

Here, the core data presents:

* `id`, the unit ID
* `alias`, the alias associated to that ID
* `type` of unit, either base or derived (see **A brief note on design and methodology** to understand the significance of this)
* `category`, the defined category within which conversions may be performed
* `srp`, the associated standard reference unit for that `category`
* `model`, a list column providing the unit's conversion model for converting to the associated `srp` unit

Together, these variables form the core operating dataset for the `unicon` functionality.

# A brief note on design and methodology

If your use of `unicon` falls firmly into the "casual" category, you can probably skip this section. However, we recommend reading it (or returning to it later) if you plan to rely on `unicon`'s functions in any kind of higher-stakes setting, or on developing/modifying your own conversion protocols. In this section, we aim to take you through some of the methods and design approach behind the `unicon` system.

## Standard reference units

All unit conversions are performed with respect to a single reference unit defined per category:

```{r simple-process}
DiagrammeR::mermaid("
graph LR;
  A[/Input Unit/] --> B(Standard Reference Unit)
  B --> C[/Output Unit/];
")
```

Within the package data, each base unit is provided with its own human-readable ID, and defined in terms of its relationship to that single reference unit. A JSON schema for each base unit encapsulates this information.

The following is the schema for the unit ID `acres`, a non-standard unit of area commonly used in agriculture:

```{r base-schema, class.output="json", comment=NA, echo=FALSE}
acre <- jsonlite::read_json(
  system.file("units", "base", "area", "acre.json", package = "unicon"),
  simplifyVector = FALSE
)
jsonlite::toJSON(acre, auto_unbox = TRUE, pretty = TRUE)
```

Within `unicon`, the standard reference unit for area is a hectare (ID: `ha`), and an acre is defined as approximately 0.4 of a hectare. If converting from acres to any other unit of area, the quantity provided will always first be converted to hectares. Likewise, if we know how to convert from acres to hectares, we know how to do the reverse. It is therefore easy to convert any other units of area to acres, provided we can first convert them to hectares; and since all units of area are defined by their relationship to a hectare, this is a given.

This approach is fundamental, and allows the package to focus on maintaining only one conversion relationship per unit. This minimises the physical size of data maintained and minimises the potential for internal inconsistencies.

## Unit aliases

The other main attribute of a unit is its `alias` list. This is a list of the human-readable names by which the unit may be called in the real world. For the unit with ID `acre`, its possible aliases are ``r as.character(acre$alias)``. Together with the standard reference unit and conversion model, these pieces of information form the core "profile" of a unit, representing the main things we need to know about it in order to convert to or from it, and to correctly recognise it in the real world.

A couple of simple transforms are used to keep the alias list as focused as possible. Firstly, aliases are lowercase and treated as case-insensitive. Secondarily, aliases are devoid of whitespace and whitespace is stripped from user-supplied inputs, a feature which becomes particularly useful when dealing with derived units (see below).

Combining all of this together with the established input --> standard reference unit --> output convention gives us the full process flow of the `unicon` function:

```{r full-process}
DiagrammeR::mermaid("
graph LR
  A1[/Input unit/]
  B1(Input unit alias)
  C1(Input unit ID)

  A2[/Output unit/]
  B2(Output unit alias)
  C2(Output unit ID)

  D(Standard Reference Unit ID)

  E[[Check for errors in unit recognition]]
  E1[/Warn of issues/]

  F[[Convert input value to SRU]]
  G[[Convert SRU value to output]]
  H[/Return outputs/]

  A1 --> B1 --> C1
  A2 --> B2 --> C2
  C1 --> D
  C2 --> D

  C1 --> E
  C2 --> E
  D --> E
  E --> E1

  D --> F --> G --> H
  C1 --> F
  C2 --> G
")
```

## Derived units

So far, we have only discussed base units with schemas directly defined in the core `unicon` package data. The `unicon` system draws a clear distinction between "base" units, like hectares and acres, and "derived" units like kilograms per hectare or gallons per acre. This distinction is designed, as with the single point of reference for base units, to minimise repetition and the need for redundant and potentially inconsistent data. It works on the principle that if we know how to convert to and from the base units which make up a derived unit, we must also know how to convert to or from that derived unit to another in its category.

For example, if we know how to convert kilograms to pounds and from hectares to acres, then by definition, we know how to convert from kilograms per hectare to pounds per acre. We don't need to record any additional information in order to achieve this.

In `unicon`, a separate JSON schema exists to define derived units. Since the heavy lifting is done by the base unit schemas, this schema can be extremely minimal, as with the below example for the derived unit with ID `rate` (mass per unit area):

```{r derived-schema, class.output="json", comment=NA, echo=FALSE}
rate <- jsonlite::read_json(
  system.file("units", "derived", "area_density.json", package = "unicon"),
  simplifyVector = FALSE
)
jsonlite::toJSON(rate, auto_unbox = TRUE, pretty = TRUE)
```
In this schema, `x` and `y` define the two base unit categories which make up the derived category, and `operator` defines the relationship between them. 

## The build process

The final tenet of the `unicon` system is to minimise the amount of computation that is done at execution time. This ensures calculation stays fast, and aids in transparency, since the number of steps to get from input to output remains minimal and identical for all units. To achieve this, the build process, which translates from raw JSON schemas to package internal data, aims to do as much pre-processing as possible.

Firstly, all base units schemas are read and their models codified.

Secondly, these base unit models, together with the derived unit schemas, are used to build models and aliases for derived units. This works as follows:

Each `operator` as defined in the derived unit schemas requires its own definition and alias list, as per this one, for division:

```{r operator-schema, class.output="json", comment=NA, echo=FALSE}
divide <- jsonlite::read_json(
  system.file("units", "operators", "divide.json", package = "unicon"),
  simplifyVector = FALSE
)
jsonlite::toJSON(divide, auto_unbox = TRUE, pretty = TRUE)
```

For each derived unit schema, aliases are generated by cross-joining:

1. All aliases for base unit `x`
2. All aliases for the defined `operator`
3. All aliases for base unit `y`

An ID for the derived unit is created by pasting together the IDs for contributing components i.e. `paste(x, operator, y)`. For example, a rate unit combining tonnes (ID: `t`) and hectares (ID: `ha`) via division (ID: `__`) would receive ID `t__ha`, with aliases `t/ha`, `tperha`, `tonnes/ha`, `tonnesperha`, and so on. It is easy to see how one simple relationship can quickly generate a bewildering number of aliases.

The final stage of this build process is to mathematically recombine the contributing models, so that the model for `t__ha` and all units in the `rate` category remain a simple single-factor slope model. Note that it is not possible to combine models which have a none-zero intercept in this way, and these are excluded from derived unit calculations.

The package datasets created are referenced by all `unicon` functions, and may also be examined/queried by the user via the `unicon_help` function, discussed above.

```{r unicon-help, echo=FALSE}
print(unicon_help(), n = 10L)
```

# Adding missing database information

Please raise an issue in the [GitHub](https://github.com/unicon/) repository and the authors will add these in. 

# Caveats

{unicon} deviates slightly away from SI standards to increase real-life usefulness: 

* For mass, the standard reference point unit is grams (g) not kilograms (kg)
* For temperature, the standard reference point unit is grams celcuis (c) not kelvin (k)
* For area, the standard reference point is hectares (ha) not metres squared (m2)
* For volume, the standard reference point is litres (l) not metres cubed (m3)

Other things to note are:

* Area and volume are considered to be base units as well as derived units
* Speed is listed as speed not velocity
* Length is listed as length not distance
* Month is **NOT** included in the package due to the variation in the number of days and weeks per month
* Year is recorded as **365.25 days** to account for leap years.
