---
title: "unicon"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(unicon)
```

# Introduction

The International System of Units (SI) standardises unit measurements, preventing different unit systems (most notably metric and imperial) being used interchangeably throughout the world to convey the same information, allowing meaningful comparisons. Simple calculations are often needed to ensure that measurements are in SI units, requiring conversion from one unit to another, for example yards to metres. If a unit of measurement is derived based on data from other units, such as speed, a calculation is required, converting the base units into SI first. This may trip up the best mathematicians or lead to annoying additions in code to solve these issues. 

The {unicon} package provides:

1. A database of *base* and *derived* units and their SI equivalents, including their various names and notations, and the equations using base units to calculate derived units.
2. Functions to convert non-SI units to SI units and vice versa.
3. Functions to derive additional derived units from existing data, for example, volume to concentration.

# Database of base and derived units

The unit.conversion package runs on three key mappings:

1. Names for units and the recognised aliases used in this package. E.g. cm, centimetres, and centimeters are names for cm.
2. Calculations for converting non-SI derived units into SI derived units. E.g. multiplying cm/s by 0.01 to achieve m/s.
3. Lists of what the recognised SI unit is. E.g. cm is m.

These data are used in the functions described below.

# Conversion of non-SI units to SI units and vice versa

## Converting to SI

## Converting from SI

# Derivation of additional units from existing data

# Adding missing database information

Please raise an issue in the [GitHub](https://github.com/unicon/) repository and the authors will add these in. 


###



The International Standard of Units (SI, from the french *'Système international d'unités'*) is the most employed unit of measurement used around the world. It has seven base units:

* Time - second (s)
* Length - metre (m)
* Mass - kilogram (kg)
* Temperature - kelvin (K)
* Amount of Substance - mole (mol)
* Electric Current - ampere (A)
* Luminous Intensity - candela (cd)


The Imperial System of Units was first developed by the British in the nineteeth century, and has wide adoption across countries previously part of the British Empire. This measurement system's base units comprise of:

* Length (e.g. yards)
* Mass (e.g. stone)
* Area (e.g. acre)
* Volume (e.g. pint)


The {unicon} ('unit conversion') package aims to provide a tool to enable users to convert quickly between different unit of measurements, either to or from SI units or to or from other units via SI units. 


# Relationships 

Length * Length = Area
where lengths can be different values 

Length * Length * Length = Volume
Or Area * Length = Volume
where lengths can be different values 

1.	Density = Mass / Volume

2.	Speed (or Velocity) = Distance / Time

3.	Acceleration = Change in Speed (speed 1 - speed 2) / Time

4.	Force = mass * Acceleration or mass * 9.81 (gravitational pull)

5.	Pressure = Force / area

6.	Concentration = mol / Volume 

7.	Volumetric Flow Rate = Area * Speed = Volume / Time 

8.	Mass Flow Rate = Density * Volumetric Flow Rate





# unicon differences to SI

* For mass, the standard is grams (g) not kilograms (kg)
* Celcuis not kelvin
* Area and Volume are considered to be base units rather than derived
* For area, the standard is hectares (ha) not metres squared (m2)
* For volume, the standard is litres (l) not metres cubed (m3)
* Speed is listed as speed not velocity
* if mass missing 9.81 is assumed for calculating acceleration or force

