---
title: "unicon"
output: rmarkdown::html_vignette
toc: true
toc_float: true
number_sections: true
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(unicon)
```

# Introduction

Different unit systems (most notably metric and imperial) are used interchangeably around the world to convey the same information. The International System of Units (SI, from the french *'Système international d'unités'*) aims to standardise unit measurements, allowing meaningful comparisons. However, the problem remains if data is collected using non-SI measurement systems. The {unicon} ('unit conversion') package provides a simple tool to enable users to quickly convert between different units of measurement using:

1. A database of base and derived units, including their various names, notations, and SI equivalents;
2. Functions to convert non-SI units to SI units and vice versa;
3. Functions to calculate additional units from existing data.

# Database of base and derived units

Three key mappings are created within the package:

1. Names and aliases for units. E.g. cm, centimetres, and centimeters are aliases for cm.
2. Calculations for converting non-SI derived units into SI derived units. E.g. multiplying cm/s by 0.01 to achieve m/s.
3. A list of what the recognised SI unit is within the {unicon} package. E.g. SI for length is m.
4. Relationships between different metrics to allow calculation of new metrics. E.g. Area * Length = Volume.

These data are used in the functions described below.

## Metric relationships 

The relationships between metrics that are included in this package are:

1.  **Length * Length = Area** where lengths can be different values. Please note *Area* is also listed as a base unit.

2.  **Area * Length = Volume** where *Area* and *Volume* are also listed as a base units. Please calculate *Area* from first equation if unknown. 

3.	**Area Density = Mass / Volume**

4.	**Speed = Length / Time**

5.	**Acceleration = Change in Speed / Time**

6.	**Force = Acceleration * Mass** where *Mass* is assumed as 9.81 if considering gravitational force

7.	**Pressure = Force / Area**

8.	**Concentration = Moles / Volume**

9.	**Volumetric Flow Rate = Area * Speed**

10. **Volumetric Flow Rate = Volume / Time**

11.	**Mass Flow Rate = Volume Density * Volumetric Flow Rate**

# Examples

To convert data into SI units:

```{r}
# e.g. miles to SI metres

unicon_full(
  value_in = 8,
  unit_in = "miles"
)
```

To calculate a non-SI unit:

```{r}
# e.g. miles to km (function does this via m)

unicon_full(
  value_in = 8,
  unit_in = "miles",
  unit_out = "km"
)
```

To calculate a new metric from existing data:

```{r}
# e.g. speed from length / time

unicon_catrel(
  parent_1_unit_in = "miles",
  parent_2_unit_in = "hour",
  parent_1_value_in = c(1, 2, 3, 4, 5, 6),
  parent_2_value_in = c(9, 8, 7, 5, 4, 2),
  unit_out = "km/day"
)

# if unit_out unspecified, will convert to SI (in this example m/s)
```

## Acceleration 

```{r}
## acceleration = change in speed / time
## i.e. (speed 1 - speed 2) / time

speed_1 <- unicon_full(
  value_in = 8,
  unit_in = "miles"
)

speed_2 <- unicon_full(
  value_in = 16,
  unit_in = "km"
)

unicon_catrel(
  parent_1_unit_in = "m/s",
  parent_2_unit_in = "h",
  parent_1_value_in = (speed_1 - speed_2),
  parent_2_value_in = 2
)
```

# Adding missing database information

Please raise an issue in the [GitHub](https://github.com/unicon/) repository and the authors will add these in. 

# Caveats

{unicon} deviates slightly away from SI standards to increase real-life usefulness: 

* For mass, the standard unit is grams (g) not kilograms (kg)
* For temperature, the standard unit is grams celcuis (c) not kelvin (k)
* For area, the standard is hectares (ha) not metres squared (m2)
* For volume, the standard is litres (l) not metres cubed (m3)

Other things to note are:

* Area and volume are considered to be base units as well as derived units
* Speed is listed as speed not velocity
* Length is listed as length not distance
* If trying to calculate force or acceleration using the assumption mass is equal to gravitational force, the value **9.81** is used
* Month is **NOT** included in the package due to the variation in the number of days and weeks per month
* Year is recorded as **365.25 days** to account for leap years.

