---
title: "unicon functions"
output: rmarkdown::html_vignette
toc: true
toc_float: true
number_sections: true
vignette: >
  %\VignetteIndexEntry{unicon_functions_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(unicon)
```

# `unicon_full`

The core user-facing function of the `unicon` package is `unicon_full`. Use it like this:

```{r unicon_full-1}
raw_values <- c(54.21, 71.24, 55.81, 11.33, 70.59)
raw_units <- c("tonnes / ha", "tons per acre", "t/ha", "kg /Hectare", "g/m2")
unit_out <- "tonnes / ha"

unicon_full(
  value_in = raw_values,
  unit_in = raw_units,
  unit_out = unit_out,
  pull = TRUE
)
```

The function uses `unicon`'s extensive library of unit aliases to match the highly varied raw input units to their corresponding unit IDs, and subsequently converts them to the specified output unit.

A number of features are built into the function to help keep conversions safe and transparent. Firstly, the user may wish to see the full calculation thread which leads to this outcome, which can be achieved by setting the `pull` argument to FALSE. This yields a tibble, which we'll `glimpse`:

```{r unicon_full-2}
unicon_full(
  value_in = raw_values,
  unit_in = raw_units,
  unit_out = unit_out,
  pull = FALSE
) |>
  dplyr::glimpse()
```

This returns the following information, representing the calculation process in rough left --> right chronology:

* `unit_in` The user-supplied `unit_in` argument.
* `unit_out` The user-supplied `unit_out` argument.
* `alias_in` The derived alias associated to the `unit_in` argument, generated by stripping whitespace and converting all characters to lowercase.
* `alias_out` The derived alias associated to the `unit_out` argument, derived similarly.
* `id_in` The ID acquired for the `alias_in` variable.
* `id_srp` The ID associated with the "SRP" (standard reference point) unit for the type class of the input and output units.
* `id_out` The ID acquired for the `alias_out` variable.
* `error_in` Logical; were there errors in processing the input unit?
* `error_srp` Logical; were there errors in deriving the SRP unit?
* `error_out` Logical; were there errors in deriving the output unit?
* `value_in` The user-supplied `value_in` argument.
* `value_srp` The values as converted to their SRP (standard reference point) unit.
* `value_out` The values converted to the `unit_out` argument, if supplied. If not supplied, the standard reference point unit will be returned and the values will be identical to `value_srp`.

To understand more about the significance of the `srp` unit, read **A brief note on design and methodology**, in the `unicon_vignette`.

All the core input arguments are vectorised and will attempt to replicate if supplied as scalar values. This is helpful, as above, in specifying a single unit of output, or in converting vectors of consistent unit type (`unicon_full(my_values, "tonnes/ha", "g/m2")`). It's also possible to omit the `unit_out` argument altogether; if omitted, the function will return values in the associated standard reference unit and will message the user to indicate that this has been done.

Finally, although `unicon` has a very large library of unit aliases, it is probable that it will be passed units that it does not recognise and important that it handles these clearly:

```{r unicon_full-3, warning=TRUE}
unicon_full(
  value_in = raw_values,
  unit_in = c("tonnes / ha", "tons per acre", "t/ha", "kg /Hectare", "nonsense per fiction"),
  unit_out = unit_out,
  pull = TRUE
)
```

As previously, setting the `pull` argument to `FALSE` will yield better information as to the nature of the error.

```{r unicon_full-4, warning=TRUE}
unicon_full(
  value_in = raw_values,
  unit_in = c("tonnes / ha", "tons per acre", "t/ha", "kg /Hectare", "nonsense per fiction"),
  unit_out = "tonnes / ha",
  pull = FALSE
)
```

# `unicon_lite`

While `unicon_full` serves as core user-facing function, this is merely a wrapper function around `unicon_lite`. It is recommended users use the `unicon_full` function for the following reasons:

- Completes data checks on inputted data and provides helpful warnings;
- Allows users to choose to only receive `value_out` only, rather than including all internal workings columns;
- Does not require users to know the unit ID for each alias used by the package in correct format needed for the package, although these can be provided if known. If they are provided, the function will skip the alias search, and performance may be improved.

Direct usage is discouraged for the reasons outlined above, but users can still implement `unicon_lite`:

```{r unicon_lite-1}
unicon_lite(
  value_in = raw_values,
  id_in = c("t__ha", "st__acre", "t__ha", "kg__ha", "g__m2"),
  id_out = "t__ha"
)
```

# `unicon_help`

To use the `unicon` functions, it may be useful to directly see what is recognised by the package. The internal package data is maintained in an optimal format for efficient computation, but the `unicon_help` function is designed to join and return this in the simplest format for interactive analysis:

```{r unicon_help}
unicon_help()
```

Here, the core data presents:

* `id`, the unit ID
* `alias`, the alias associated to that ID
* `type` of unit, either base or derived (see **A brief note on design and methodology** to understand the significance of this)
* `category`, the defined category within which conversions may be performed
* `srp`, the associated standard reference unit for that `category`
* `model`, a list column providing the unit's conversion model for converting to the associated `srp` unit

Together, these variables form the core operating dataset for the `unicon` functionality.

# unicon_own

The authors recognise that every unit conversion that users may want to use this package for may not exist within the package data. As such, `unicon_own` allows users to add their own base, derived, and operator data (please refer to `unicon_vignette.Rmd` for more information on how these data differ) and take advantage of the same functionality that builds the out of the box package data. Before users engage, it is recommended that the tibble produced by `unicon_help` be studied, as well as the `.json` files in the `inst` folder, to understand the data structure required to ensure correct functionality. 

`unicon_own` is actually a wrapper function (with many arguments!) which does several things:

- Formats the user's own data in the same format as existing package data using `unicon_make_own_*_data` functions
- Binds data to the existing package data, stored as `.json` files in the `inst` folder using `unicon_make_*_data_from_jsons` functions
- Joins base, derived, and operator datasets together using `unicon_join_datasets`
- Creates required mappings for use in `unicon_full` (and `unicon_lite`) using `unicon_make_unit_*` functions
- Updates the package data automatically by altering `.unicon_state`

The three datasets are built and stored internally; the active state can be checked with `unicon_own_status()` and the package data can be queried via `unicon_help()` after calling `unicon_own()`.

Users can add to base, derived, and operator data at the same time, or just one dataset, as all arguments in `unicon_own` default to `NA`. It is hypothesised that users will mostly add their own base data, with derived data being rarely needed to be added. The author's don't envisage a use for other operators apart from multiply and divide, but please get in touch if you disagree with us!

An example of use is as follows: Users could have a defined 'package' size of 12.5 kg or 6.25 kg. Users may have information about the number of packages of certain size or the weight in kg of the packages, but not an easy way to convert from one to the other.

```{r}
lp <- "lp"
sp <- "sp"

unicon_own(
  base_id    = c(rep(lp, 2), rep(sp, 2)),
  base_alias = c("lp", "large package", "sp", "small package"),
  category   = c(rep("mass", 4)),
  srp        = c(rep("g", 4)),
  slope      = c(rep(12500, 2), rep(6250, 2)),
  intercept  = c(rep(0, 4)) # default; shown for clarity
)

unicon_own_status()

unicon_help() |>
  dplyr::filter(id %in% c("lp", "sp"))
```

Then use `unicon_full` as usual:

```{r}
unicon_full(
  value_in = c(3, 5),
  unit_in = c("large package", "small package"),
  unit_out = "kg"
)
```

Authors also suggest that users could use this functionality for money exchange rates between different currencies, where the exchange rate can easily be updated.

If users want to revert back to the package data, all they need to do is:

```{r}
unicon_reset_units()
```

# `unicon_advance`

The `unicon_advance` function is for the hard-core unit converters out there - those that have two metrics and want to create a third metric in the desired unit.

The metric relationships that are stored in the `relationships` package dataset are:

```{r}
unicon:::relationships
```

The beady eyed user will see that the same relationship is repeated four times, so that every combination of `id`, `x`, `y`, and `operator` is included.

Additionally, if users add their own derived data, this will automatically populate the above dataset for use in their workflows.

This functionality works by:
- Identifying the metric category type for the supplied x and y `unit_in` e.g. kg for weight
- Identifying the relationship between x and y
- Calculating the desired metric first in the package's defined SRP for that metric, before using the package's unit conversion functionality to calculate the desired unit for that metric

As an example to how calculate a new metric from existing data is speed from length / time. If `unit_out` is unspecified, the function will convert to the package's SRP, in this example m/s.

```{r unicorn_advance}
unicon_advance(
  x_unit_in = rep("miles", 6),
  y_unit_in = rep("hour", 6),
  x_value_in = c(1, 2, 3, 4, 5, 6),
  y_value_in = c(9, 8, 7, 5, 4, 2),
  unit_out = rep("km/day", 6)
)
```

Like `unicon_full`, `unicon_advance` has a `pull` argument, and it works in the same way; if set to `TRUE` (default), only the `value_out` is returned with a reduced warning. If set to `FALSE` a tibble is returned which shows all the `unicon_full` functionality for x, y, and the desired unit (we did warn you). This is useful in trouble shooting should an error in data input occur. 
